---
title: "Differential Gene Expression Analysis: AD vs. Control (GSE67333)"
author: "KASHVI SHAH"
date: "`2025-08-28`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

## Introduction

This document outlines the steps for performing a differential gene expression (DGE) analysis using RNA-Seq count data. The goal is to identify genes that are differentially expressed between Alzheimer's Disease (AD) samples and Control samples from GSE67333. The analysis pipeline uses the **limma-voom** workflow, which is well-suited for this type of data.

The main steps are: 1. Data loading and preparation. 2. Filtering low-expression genes and data normalization. 3. Differential expression analysis using `limma` and `voom`. 4. Saving the results.

------------------------------------------------------------------------

## 1. Environment Setup

### A. Project-relative paths

All inputs and outputs are resolved from the repository root with `here::here()`.

```{r paths}
dataset_id <- "GSE67333"
input_file <- here::here("data", "processed", dataset_id, "Edited_raw_counts.csv")
outdir <- here::here("results", "differential_expression", dataset_id)
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
result_file <- function(filename) file.path(outdir, filename)
```

### B. Load Required Libraries

We load the R packages required for the analysis. \* **edgeR**: Used for creating the `DGEList` object, filtering, and normalization. \* **limma**: The core package for differential expression analysis using linear models. It includes the `voom` function. \* **ggplot2**: A powerful library for creating visualizations, such as boxplots. \* **reshape2**: Used for data manipulation, specifically the `melt` function to prepare data for `ggplot2`.

```{r load_libraries}
library(edgeR)
library(limma)
library(ggplot2)
library(reshape2)
library(here)
```

------------------------------------------------------------------------

## 2. Data Loading and Preparation

### A. Load Raw Count Data

We load the raw gene counts from a CSV file. \* `check.names = FALSE` prevents R from converting special characters in column names (e.g., hyphens). \* `stringsAsFactors = FALSE` ensures character columns are not converted to factors.

```{r load_data}
counts_raw <- read.csv(input_file, check.names = FALSE, stringsAsFactors = FALSE)
```

### B. Clean and Format Count Data

The raw data needs some cleaning before it can be used. 1. **Rename Gene Column**: Ensure the first column is named "Gene". 2. **Clean Sample Names**: Remove the leading "X" that R sometimes adds to column names that start with a number. 3. **Handle Duplicate Genes**: `make.unique` is used to ensure all gene names (which will become row names) are unique. This is crucial for creating the count matrix. 4. **Check for Missing Values (NA)**: `sum(is.na(counts_raw))` is used to check any missing counts in dataset and replace it with 0 if detected. 5. **Create Numeric Matrix**: The data is converted into a numeric matrix, with gene names as row names and sample names as column names. This is the required format for `edgeR`.

```{r clean_counts}
# Ensure first column is "Gene" and clean colnames (remove leading X)
colnames(counts_raw)[1] <- "Gene"
colnames(counts_raw) <- sub("^X", "", colnames(counts_raw))

# Check and handle potential duplicate gene names
counts_raw[duplicated(counts_raw[, 1]), 1]
counts_raw$Gene <- make.unique(as.character(counts_raw$Gene))
rownames(counts_raw) <- counts_raw$Gene

# Check and handle potential no NA values in the dataset
sum(is.na(counts_raw))
counts_raw[is.na(counts_raw)] <- 0

# Build numeric counts matrix
counts_mat <- counts_raw[, -1, drop = FALSE]
counts_mat <- as.matrix(sapply(counts_mat, as.numeric))
rownames(counts_mat) <- counts_raw$Gene
```

### C. Create Metadata

We create a metadata data frame that contains information about each sample. The condition ("AD" or "Control") is determined from the sample names (e.g., names starting with "LOAD" are AD, "CTRL" are Control). This metadata is then aligned with the columns of our count matrix.

```{r create_metadata}
sample_names <- colnames(counts_mat)
condition <- ifelse(
  grepl("^LOAD", sample_names), "AD",
  ifelse(grepl("^CTRL", sample_names), "Control", NA)
)

metadata <- data.frame(
  Sample = sample_names,
  Condition = factor(condition, levels = c("Control", "AD"))
)

# Drop non-matching (if any) and align
keep_samps <- !is.na(metadata$Condition)
metadata   <- metadata[keep_samps, , drop = FALSE]
counts_mat <- counts_mat[, metadata$Sample, drop = FALSE]

cat("Samples per group:\n")
print(table(metadata$Condition))
```

------------------------------------------------------------------------

## 3. Data Quality Assessment and Filtering

### A. Boxplot of Raw Counts (Before Filtering)

To assess the distribution of counts across samples, we generate a boxplot. This helps identify any major outliers or batch effects. We use log2-transformed counts (`log2(count + 1)`) for better visualization.

```{r boxplot_before}
log_counts <- log2(counts_mat + 1)
log_counts_df <- as.data.frame(log_counts)
log_counts_df$Gene <- rownames(log_counts_df)
melted_counts <- melt(log_counts_df, id.vars = "Gene",
                      variable.name = "Sample", value.name = "LogCount")

ggplot(melted_counts, aes(x = Sample, y = LogCount)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(Count + 1)") +
  ggtitle("Boxplot of Raw Log-Counts (Before Filtering)")

ggsave(result_file("_boxplot_raw_logcounts.png"), width = 10, height = 5, dpi = 300)
```

### B. Create DGEList, Normalize, and Filter

Now, we prepare the data for differential expression analysis. 1. **DGEList Object**: We create a `DGEList` object from the count matrix and metadata. This is the main data object used by `edgeR`. 2. **Normalization**: We apply TMM (Trimmed Mean of M-values) normalization using `calcNormFactors`. This method adjusts for differences in library size and composition between samples. 3. **Filtering**: We filter out genes with very low expression. These genes have little statistical power and can add noise to the analysis. Here, we keep genes that have a log2-CPM (Counts Per Million) value greater than 1 in at least the minimum number of samples in any condition.

```{r filter_normalize}
dge <- DGEList(counts = counts_mat, samples = metadata, group = metadata$Condition)
dge <- calcNormFactors(dge, method = "TMM")
cat(sprintf("Kept genes before filtering: %d\n", nrow(dge)))

# Define filtering threshold
log2cpm_raw <- cpm(dge, log = TRUE, prior.count = 1)
min_n <- min(table(metadata$Condition))
keep <- rowSums(log2cpm_raw > 1) >= min_n

# Filter the DGEList and recalculate norm factors
dge_filt <- dge[keep, , keep.lib.sizes = FALSE]
dge_filt <- calcNormFactors(dge_filt, method = "TMM")

cat(sprintf("Kept genes after filtering: %d\n", nrow(dge_filt)))
```

### C. Boxplot of Filtered Counts (After Filtering)

We create another boxplot to visualize the data after filtering and normalization. The distributions between samples should now be more aligned. The y-axis now represents Log2-CPM, reflecting the normalization.

```{r boxplot_after}
# Recalculate the log2-CPM on the filtered DGEList object.
log2cpm_filtered <- cpm(dge_filt, log = TRUE, prior.count = 1)

log_counts_df_after <- as.data.frame(log2cpm_filtered)
log_counts_df_after$Gene <- rownames(log_counts_df_after)
melted_counts_after <- melt(log_counts_df_after, id.vars = "Gene",
                            variable.name = "Sample", value.name = "LogCount")

ggplot(melted_counts_after, aes(x = Sample, y = LogCount)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(CPM + 1)") +
  ggtitle("Boxplot of Filtered & Normalized Log-CPM")

ggsave(result_file("_boxplot_filtered_logcounts.png"), width = 10, height = 5, dpi = 300)

write.csv(log2cpm_filtered,
          file = result_file("log2cpm_filtered_TMM.csv"),
          row.names = TRUE)
```

------------------------------------------------------------------------

## 4. Differential Expression Analysis with limma-voom

### A. Define the Design Matrix

We create a design matrix using the `model.matrix` function. This matrix specifies the model that will be fit to each gene. The formula `~ Condition` models the gene expression based on the `Condition` variable (AD vs. Control), with "Control" as the reference level.

```{r design_matrix}
design <- model.matrix(~ Condition, data = metadata)
rownames(design) <- colnames(dge_filt)
head(design)
```

### B. Run Voom and Fit the Linear Model

The `limma-voom` workflow is ideal for RNA-Seq data. 1. **`voom`**: This function transforms the count data to log2-CPM, estimates the mean-variance relationship, and computes precision weights for each observation. These weights are then used in the linear model fitting, allowing `limma` to handle count data appropriately. The plot shows the relationship between expression level (x-axis) and variance (y-axis). 2. **`lmFit`**: Fits a linear model for each gene based on the provided design matrix. 3. **`eBayes`**: Applies empirical Bayes moderation to the standard errors, borrowing information across all genes to obtain more stable and powerful statistical estimates.

```{r voom_limma, fig.cap="Voom Mean-Variance Plot"}
v <- voom(dge_filt, design, plot = TRUE)
fit <- lmFit(v, design)
fit <- eBayes(fit)
```

### C. Extract and Subset DEGs

We extract the results using `topTable`. \* `coef = "ConditionAD"` specifies that we are interested in the coefficient for the AD group, which represents the log2 fold change of AD relative to Control. \* `number = Inf` ensures all genes are returned. \* We then subset these results to find **significant** DEGs using a log-fold-change (logFC) threshold of `> 0.58` (equivalent to a fold change of \~1.5) and a P-value threshold of `< 0.05`.

```{r extract_degs}
deg <- topTable(fit, coef = "ConditionAD", number = Inf, sort.by = "P")
deg_sig_adjusted <- subset(deg, abs(logFC) > 0.58 & adj.P.Val < 0.05)
deg_sig_unadjusted <- subset(deg, abs(logFC) > 0.58 & P.Value < 0.05)
# Primary filtered set for figures, gene lists, and enrichment.
deg_sig <- deg_sig_unadjusted

head(deg_sig)
cat(sprintf("\nTotal significant DEGs found: %d\n", nrow(deg_sig)))
cat(sprintf("Adjusted-p-value DEGs: %d; unadjusted-p-value DEGs: %d\n",
            nrow(deg_sig_adjusted), nrow(deg_sig_unadjusted)))
```

------------------------------------------------------------------------

## 5. Save Results

Finally, we save the lists of significant DEGs. 1. A CSV file containing the full statistics for all significant DEGs. 2. Two separate text files containing only the gene names for upregulated and downregulated genes, respectively. This format is useful for downstream pathway or gene set enrichment analysis.

```{r save_results}
# Save all significant DEGs
write.csv(deg, result_file("full_DE_results.csv"), row.names = TRUE)
write.csv(deg_sig, result_file("significant_DEGs.csv"), row.names = TRUE)
write.csv(deg_sig_adjusted, result_file("significant_DEGs_adjusted.csv"), row.names = TRUE)
write.csv(deg_sig_unadjusted, result_file("significant_DEGs_unadjusted.csv"), row.names = TRUE)

# Save lists of up- and down-regulated gene names
up_genes <- rownames(deg_sig[deg_sig$logFC > 0, , drop = FALSE])
down_genes <- rownames(deg_sig[deg_sig$logFC < 0, , drop = FALSE])

write.table(up_genes, result_file("upregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)
write.table(down_genes, result_file("downregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)

cat(sprintf("Found %d upregulated and %d downregulated genes.\n", length(up_genes), length(down_genes)))

# (Optional) Look at top 20 up/down regulated genes
top20_up <- head(deg_sig[deg_sig$logFC > 0, ][order(deg_sig$P.Value[deg_sig$logFC > 0]), ], 20)
top2 <- head(deg_sig[deg_sig$logFC < 0, ][order(deg_sig$P.Value[deg_sig$logFC < 0]), ], 20)
```
```{r}
library(pheatmap)

# Enrichment
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

# pick top N significant DEGs

topN <- 20
top_genes <- rownames(deg_sig)[order(deg_sig$P.Value)][1:min(topN, nrow(deg_sig))]

# use voom normalized expression

expr_mat <- v$E[top_genes, metadata$Sample, drop = FALSE]

# Z-score per gene for visualization

expr_z <- t(scale(t(expr_mat)))
expr_z[is.na(expr_z)] <- 0

# sample annotation

ann_col <- data.frame(Condition = metadata$Condition)
rownames(ann_col) <- metadata$Sample
png(result_file("_heatmap_top20_genes.png"), width = 4000, height = 2400, res = 600)
pheatmap(expr_z,
annotation_col = ann_col,
show_colnames = FALSE,
fontsize_row = 7,
main = paste0("Heatmap Top ", length(top_genes)))

dev.off()
```

```{r}
# assumes rownames are gene SYMBOLs; if they are ENSG IDs, tell me and I'll swap mapping

sig_symbols <- rownames(deg_sig)

conv <- bitr(sig_symbols,
fromType = "SYMBOL",
toType   = c("ENTREZID","ENSEMBL"),
OrgDb    = org.Hs.eg.db)

# keep unique ENTREZ

sig_entrez <- unique(conv$ENTREZID)

# up/down sets (optional separate enrichment)

up_symbols   <- rownames(deg_sig[deg_sig$logFC > 0, , drop=FALSE])
down_symbols <- rownames(deg_sig[deg_sig$logFC < 0, , drop=FALSE])

conv_up <- bitr(up_symbols, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
conv_dn <- bitr(down_symbols, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

up_entrez <- unique(conv_up$ENTREZID)
dn_entrez <- unique(conv_dn$ENTREZID)

cat("Mapped significant genes to Entrez:", length(sig_entrez), "\n")
cat("Up:", length(up_entrez), "Down:", length(dn_entrez), "\n")
```

```{r}
ego_bp <- enrichGO(gene          = dn_entrez,
OrgDb         = org.Hs.eg.db,
keyType       = "ENTREZID",
ont           = "BP",
pAdjustMethod = "BH",
qvalueCutoff  = 0.05,
readable      = TRUE)

ego_cc <- enrichGO(gene = dn_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="CC", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

ego_mf <- enrichGO(gene = dn_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="MF", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

# Save tables

write.csv(as.data.frame(ego_bp), result_file("GSE6733_GO_BP_dn.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_cc), result_file("GSE6733_GO_CC_dn.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_mf), result_file("GSE6733_GO_MF_dn.csv"), row.names = FALSE)

# Plot

p1 <- dotplot(ego_bp, showCategory = 20) + ggtitle("GO Biological Process (ORA)")
p2 <- dotplot(ego_cc, showCategory = 15) + ggtitle("GO Cellular Component (ORA)")
p3 <- dotplot(ego_mf, showCategory = 15) + ggtitle("GO Molecular Function (ORA)")

print(p1); print(p2); print(p3)

ggsave(result_file("_GO_BP_dotplot.png"), p1, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_CC_dotplot.png"), p2, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_MF_dotplot.png"), p3, width = 9, height = 6, dpi = 300)
```

```{r}
ekegg_up <- enrichKEGG(gene=up_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

ekegg_dn <- enrichKEGG(gene=dn_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

# Convert Entrez IDs in results to gene symbols for readability
ekegg_up <- setReadable(ekegg_up, OrgDb=org.Hs.eg.db, keyType="ENTREZID")
ekegg_dn <- setReadable(ekegg_dn, OrgDb=org.Hs.eg.db, keyType="ENTREZID")

write.csv(as.data.frame(ekegg_up), result_file("KEGG_UP.csv"), row.names=FALSE)
write.csv(as.data.frame(ekegg_dn), result_file("KEGG_DOWN.csv"), row.names=FALSE)

p_kegg_up <- dotplot(ekegg_up, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Upregulated)")
p_kegg_dn <- dotplot(ekegg_dn, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Downregulated)")

print(p_kegg_up); print(p_kegg_dn)

ggsave(result_file("_KEGG_up_dotplot.png"), p_kegg_up, width=9, height=10, dpi=300)
ggsave(result_file("_KEGG_down_dotplot.png"), p_kegg_dn, width=9, height=10, dpi=300)

```

```{r}
library(ggrepel)
# thresholds (match your DEG definition)
lfc_cut <- 0.58
p_cut <- 0.05

# Build plotting DF from limma results
vol <- deg
vol$Gene <- rownames(vol)

# Use adjusted p-value; protect against zeros for -log10
vol$adj.P.Val[vol$adj.P.Val == 0] <- .Machine$double.xmin

vol$Sig <- "NotSig"
vol$Sig[vol$logFC >  lfc_cut & vol$P.Value < p_cut] <- "Up"
vol$Sig[vol$logFC < -lfc_cut & vol$P.Value < p_cut] <- "Down"

# Pick top genes to label (top by FDR within Up/Down)
top_n <- 10  # change to 5–15 as you like

top_up <- vol[vol$Sig == "Up", ]
top_up <- top_up[order(top_up$P.Value), , drop=FALSE]
top_up <- head(top_up, min(top_n, nrow(top_up)))

top_dn <- vol[vol$Sig == "Down", ]
top_dn <- top_dn[order(top_dn$P.Value), , drop=FALSE]
top_dn <- head(top_dn, min(top_n, nrow(top_dn)))

label_df <- rbind(top_up, top_dn)

# Volcano plot
p_vol <- ggplot(vol, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Sig), alpha = 0.7, size = 1.6) +
  geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
  geom_hline(yintercept = -log10(p_cut), linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = Gene),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.35,
    point.padding = 0.2
  ) +
  theme_bw() +
  labs(
    title = "Volcano Plot: AD vs Control",
    x = "log2 Fold Change (AD vs Control)",
    y = "-log10(FDR)"
  )

print(p_vol)
ggsave(result_file("_volcano_AD_vs_Control.png"), p_vol, width = 8, height = 6, dpi = 300)

# Quick counts
cat("Significant Up:", sum(vol$Sig == "Up"), "\n")
cat("Significant Down:", sum(vol$Sig == "Down"), "\n")
```

## Session information

```{r session_info}
writeLines(capture.output(sessionInfo()), result_file("sessionInfo.txt"))
sessionInfo()
```
