---
title: "Differential Gene Expression Analysis: AD vs. Control (GSE95587)"
author: "Analysis Pipeline"
date: "`2025-09-03`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This document outlines the steps for performing a differential gene expression (DGE) analysis using RNA-Seq count data. The goal is to identify genes that are differentially expressed between Alzheimer's Disease (AD) samples and Healthy Control (HC) samples from the **GSE95587** dataset. The analysis pipeline uses the **limma-voom** workflow, which is well-suited for this type of data.

The main steps are:
1.  Data loading and preparation.
2.  Filtering low-expression genes and data normalization.
3.  Differential expression analysis using `limma` and `voom`.
4.  Saving the results.

***

## 1. Environment Setup

### A. Project-relative paths

All inputs and outputs are resolved from the repository root with `here::here()`.

```{r paths}
dataset_id <- "GSE95587"
input_file <- here::here("data", "processed", dataset_id, "Edited_raw_counts.csv")
outdir <- here::here("results", "differential_expression", dataset_id)
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
result_file <- function(filename) file.path(outdir, filename)
```

### B. Load Required Libraries

We load the R packages required for the analysis.
* **edgeR**: Used for creating the `DGEList` object, filtering, and normalization.
* **limma**: The core package for differential expression analysis using linear models, including the `voom` function.
* **ggplot2**: A powerful library for creating visualizations, such as boxplots.
* **reshape2**: Used for data manipulation, specifically the `melt` function to prepare data for `ggplot2`.

```{r load_libraries}
library(edgeR)
library(limma)
library(ggplot2)
library(reshape2)
library(here)
```

***

## 2. Data Loading and Preparation

### A. Load Raw Count Data

We load the raw gene counts from a CSV file.
* `check.names = FALSE` prevents R from converting special characters in column names (e.g., hyphens).
* `stringsAsFactors = FALSE` ensures character columns are not converted to factors.

```{r load_data}
counts_raw <- read.csv(input_file, check.names = FALSE, stringsAsFactors = FALSE)
```

### B. Clean and Prepare Count Data

Raw data often requires cleaning before analysis.
1.  **Format Column Names**: We ensure the identifier column is named "Gene" and remove any leading `X` characters that R might add to sample names.
2.  **Validate and Clean Data**: We check for duplicate gene names and missing values (`NA`). For this dataset, duplicates were found (e.g., "TRNAV-CAC") and resolved using `make.unique()` to ensure all gene names are unique. A check for missing values showed that there were no `NA`s in the dataset.
3.  **Create Numeric Matrix**: The cleaned data frame is converted into a numeric matrix with unique gene names as row names. This format is required for `edgeR`.

```{r clean_counts}
# Ensure first column is "Gene" and clean colnames
colnames(counts_raw)[1] <- "Gene"
colnames(counts_raw) <- sub("^X", "", colnames(counts_raw))

# --- Data Validation and Cleaning ---
# Check for duplicates. The output shows that duplicates were present.
# counts_raw[duplicated(counts_raw[,1]), 1]

# Check for NA values. The output '0' indicates no NAs were found.
# sum(is.na(counts_raw))

# Fix duplicates by making gene names unique
counts_raw$Gene <- make.unique(as.character(counts_raw$Gene))
rownames(counts_raw) <- counts_raw$Gene

# Build numeric counts matrix with genes as row names
counts_mat <- counts_raw[, -1, drop = FALSE]
counts_mat <- as.matrix(sapply(counts_mat, as.numeric))
rownames(counts_mat) <- counts_raw$Gene
```

### C. Create Metadata

We create a metadata data frame that contains information about each sample. The condition ("AD" or "Control") is determined from the sample names (e.g., names starting with "AD" are Alzheimer's, "HC" are Healthy Controls). This metadata is then aligned with the columns of our count matrix.

```{r create_metadata}
sample_names <- colnames(counts_mat)
condition <- ifelse(
  grepl("^AD", sample_names), "AD",
  ifelse(grepl("^HC", sample_names), "Control", NA)
)

metadata <- data.frame(
  Sample = sample_names,
  Condition = factor(condition, levels = c("Control", "AD"))
)

# Drop non-matching samples (if any) and align data
keep_samps <- !is.na(metadata$Condition)
metadata   <- metadata[keep_samps, , drop = FALSE]
counts_mat <- counts_mat[, metadata$Sample, drop = FALSE]

cat("Samples per group:\n")
print(table(metadata$Condition))
```

***

## 3. Data Quality Assessment and Filtering

### A. Boxplot of Raw Counts (Before Filtering)

To assess the distribution of counts across samples, we generate a boxplot. This helps identify any major outliers or batch effects. We use log2-transformed counts (`log2(count + 1)`) for better visualization.

```{r boxplot_before}
log_counts <- log2(counts_mat + 1)
log_counts_df <- as.data.frame(log_counts)
log_counts_df$Gene <- rownames(log_counts_df)
melted_counts <- melt(log_counts_df, id.vars = "Gene",
                      variable.name = "Sample", value.name = "LogCount")

ggplot(melted_counts, aes(x = Sample, y = LogCount)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(Count + 1)") +
  ggtitle("Boxplot of Raw Log-Counts (Before Filtering)")

ggsave(result_file("_boxplot_raw_logcounts.png"), width = 10, height = 5, dpi = 300)
```

### B. Create DGEList, Normalize, and Filter

Now, we prepare the data for differential expression analysis.
1.  **DGEList Object**: We create a `DGEList` object, the main data container used by `edgeR`.
2.  **Normalization**: We apply TMM (Trimmed Mean of M-values) normalization to adjust for differences in library size between samples.
3.  **Filtering**: We filter out genes with low expression that provide little statistical power. We keep genes with a log2-CPM (Counts Per Million) value greater than 1 in at least the minimum number of samples in any condition.

```{r filter_normalize}
dge <- DGEList(counts = counts_mat, samples = metadata, group = metadata$Condition)
dge <- calcNormFactors(dge, method = "TMM")
cat(sprintf("Kept genes before filtering: %d\n", nrow(dge)))

# Define filtering threshold
log2cpm_raw <- cpm(dge, log = TRUE, prior.count = 1)
min_n <- min(table(metadata$Condition))
keep <- rowSums(log2cpm_raw > 1) >= min_n

# Filter the DGEList and recalculate norm factors
dge_filt <- dge[keep, , keep.lib.sizes = FALSE]
dge_filt <- calcNormFactors(dge_filt, method = "TMM")

cat(sprintf("Kept genes after filtering: %d\n", nrow(dge_filt)))
```

### C. Boxplot of Filtered Counts (After Filtering)

We create another boxplot to visualize the data after filtering and normalization. The distributions between samples should now be more aligned.

```{r boxplot_after}
log2cpm_filtered <- cpm(dge_filt, log = TRUE, prior.count = 1)

log_counts_df_after <- as.data.frame(log2cpm_filtered)
log_counts_df_after$Gene <- rownames(log_counts_df_after)
melted_counts_after <- melt(log_counts_df_after, id.vars = "Gene",
                            variable.name = "Sample", value.name = "LogCount")

ggplot(melted_counts_after, aes(x = Sample, y = LogCount)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(CPM + 1)") +
  ggtitle("Boxplot of Filtered & Normalized Log-CPM")

ggsave(result_file("_boxplot_filtered_logcounts.png"), width = 10, height = 5, dpi = 300)

write.csv(log2cpm_filtered,
          file = result_file("log2cpm_filtered_TMM.csv"),
          row.names = TRUE)
```

***

## 4. Differential Expression Analysis with limma-voom

### A. Define the Design Matrix

We create a design matrix that specifies the model to be fit to each gene. The formula `~ Condition` models the gene expression based on the `Condition` variable (AD vs. Control), with "Control" as the reference level.

```{r design_matrix}
design <- model.matrix(~ Condition, data = metadata)
rownames(design) <- colnames(dge_filt)
head(design)
```

### B. Run Voom and Fit the Linear Model

The `limma-voom` workflow transforms the count data for linear modeling.
1.  **`voom`**: Transforms the count data to log2-CPM, estimates the mean-variance relationship, and computes precision weights for each observation.
2.  **`lmFit`**: Fits a linear model for each gene based on the design matrix.
3.  **`eBayes`**: Applies empirical Bayes moderation to the standard errors, borrowing information across genes to improve statistical power.

```{r voom_limma, fig.cap="Voom Mean-Variance Plot"}
v <- voom(dge_filt, design, plot = TRUE)
fit <- lmFit(v, design)
fit <- eBayes(fit)
```

### C. Extract and Subset DEGs

We extract the results using `topTable` and subset them to find statistically significant differentially expressed genes (DEGs). We define significance with a log-fold-change (logFC) threshold of `> 0.58` (a fold change of ~1.5) and an adjusted P-value threshold of `< 0.05`.

```{r extract_degs}
deg <- topTable(fit, coef = "ConditionAD", number = Inf, sort.by = "P")
deg_sig_adjusted <- subset(deg, abs(logFC) > 0.58 & adj.P.Val < 0.05)
deg_sig_unadjusted <- subset(deg, abs(logFC) > 0.58 & P.Value < 0.05)
# Primary filtered set for figures, gene lists, and enrichment.
deg_sig <- deg_sig_unadjusted

head(deg_sig)
cat(sprintf("\nTotal significant DEGs found: %d\n", nrow(deg_sig)))
cat(sprintf("Adjusted-p-value DEGs: %d; unadjusted-p-value DEGs: %d\n",
            nrow(deg_sig_adjusted), nrow(deg_sig_unadjusted)))
```

***

## 5. Save Results

Finally, we save the lists of significant DEGs.
1.  A CSV file containing the full statistics for all significant DEGs.
2.  Two separate text files containing only the gene names for upregulated and downregulated genes, which is useful for downstream pathway analysis.

```{r save_results}
# Save all significant DEGs
write.csv(deg, result_file("full_DE_results.csv"), row.names = TRUE)
write.csv(deg_sig, result_file("significant_DEGs.csv"), row.names = TRUE)
write.csv(deg_sig_adjusted, result_file("significant_DEGs_adjusted.csv"), row.names = TRUE)
write.csv(deg_sig_unadjusted, result_file("significant_DEGs_unadjusted.csv"), row.names = TRUE)

# Save lists of up- and down-regulated gene names
up_genes <- rownames(deg_sig[deg_sig$logFC > 0, , drop = FALSE])
down_genes <- rownames(deg_sig[deg_sig$logFC < 0, , drop = FALSE])

write.table(up_genes, result_file("upregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)
write.table(down_genes, result_file("downregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)

cat(sprintf("Found %d upregulated and %d downregulated genes.\n", length(up_genes), length(down_genes)))

# (Optional) Look at top 20 up/down regulated genes
top20_up <- head(deg_sig[deg_sig$logFC > 0, ][order(deg_sig$P.Value[deg_sig$logFC > 0]), ], 20)
top20_down <- head(deg_sig[deg_sig$logFC < 0, ][order(deg_sig$P.Value[deg_sig$logFC < 0]), ], 20)
```

```{r}
library(pheatmap)

# Enrichment
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

# pick top N significant DEGs

topN <- 20
top_genes <- rownames(deg_sig)[order(deg_sig$P.Value)][1:min(topN, nrow(deg_sig))]

# use voom normalized expression

expr_mat <- v$E[top_genes, metadata$Sample, drop = FALSE]

# Z-score per gene for visualization

expr_z <- t(scale(t(expr_mat)))
expr_z[is.na(expr_z)] <- 0

# sample annotation

ann_col <- data.frame(Condition = metadata$Condition)
rownames(ann_col) <- metadata$Sample
png(result_file("_heatmap_top20_genes.png"), width = 4000, height = 2400, res = 600)
pheatmap(expr_z,
annotation_col = ann_col,
show_colnames = FALSE,
fontsize_row = 7,
main = paste0("Heatmap Top ", length(top_genes)))

dev.off()
```

```{r}
# assumes rownames are gene SYMBOLs; if they are ENSG IDs, tell me and I'll swap mapping

sig_symbols <- rownames(deg_sig)

conv <- bitr(sig_symbols,
fromType = "SYMBOL",
toType   = c("ENTREZID","ENSEMBL"),
OrgDb    = org.Hs.eg.db)

# keep unique ENTREZ

sig_entrez <- unique(conv$ENTREZID)

# up/down sets (optional separate enrichment)

up_symbols   <- rownames(deg_sig[deg_sig$logFC > 0, , drop=FALSE])
down_symbols <- rownames(deg_sig[deg_sig$logFC < 0, , drop=FALSE])

conv_up <- bitr(up_symbols, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
conv_dn <- bitr(down_symbols, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

up_entrez <- unique(conv_up$ENTREZID)
dn_entrez <- unique(conv_dn$ENTREZID)

cat("Mapped significant genes to Entrez:", length(sig_entrez), "\n")
cat("Up:", length(up_entrez), "Down:", length(dn_entrez), "\n")
```

```{r}
ego_bp <- enrichGO(gene          = dn_entrez,
OrgDb         = org.Hs.eg.db,
keyType       = "ENTREZID",
ont           = "BP",
pAdjustMethod = "BH",
qvalueCutoff  = 0.05,
readable      = TRUE)

ego_cc <- enrichGO(gene = dn_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="CC", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

ego_mf <- enrichGO(gene = dn_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="MF", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

# Save tables

write.csv(as.data.frame(ego_bp), result_file("GSE95587_GO_BP_dn.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_cc), result_file("GSE95587_GO_CC_dn.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_mf), result_file("GSE95587_GO_MF_dn.csv"), row.names = FALSE)

# Plot

p1 <- dotplot(ego_bp, showCategory = 20) + ggtitle("GO Biological Process (ORA)")
p2 <- dotplot(ego_cc, showCategory = 15) + ggtitle("GO Cellular Component (ORA)")
p3 <- dotplot(ego_mf, showCategory = 15) + ggtitle("GO Molecular Function (ORA)")

print(p1); print(p2); print(p3)

ggsave(result_file("_GO_BP_dotplot.png"), p1, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_CC_dotplot.png"), p2, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_MF_dotplot.png"), p3, width = 9, height = 6, dpi = 300)
```

```{r}
ekegg_up <- enrichKEGG(gene=up_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

ekegg_dn <- enrichKEGG(gene=dn_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

# Convert Entrez IDs in results to gene symbols for readability
ekegg_up <- setReadable(ekegg_up, OrgDb=org.Hs.eg.db, keyType="ENTREZID")
ekegg_dn <- setReadable(ekegg_dn, OrgDb=org.Hs.eg.db, keyType="ENTREZID")

write.csv(as.data.frame(ekegg_up), result_file("KEGG_UP.csv"), row.names=FALSE)
write.csv(as.data.frame(ekegg_dn), result_file("KEGG_DOWN.csv"), row.names=FALSE)

p_kegg_up <- dotplot(ekegg_up, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Upregulated)")
p_kegg_dn <- dotplot(ekegg_dn, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Downregulated)")

print(p_kegg_up); print(p_kegg_dn)

ggsave(result_file("_KEGG_up_dotplot.png"), p_kegg_up, width=9, height=10, dpi=300)
ggsave(result_file("_KEGG_down_dotplot.png"), p_kegg_dn, width=9, height=10, dpi=300)

```

```{r}
library(ggrepel)
# thresholds (match your DEG definition)
lfc_cut <- 0.58
p_cut <- 0.05

# Build plotting DF from limma results
vol <- deg
vol$Gene <- rownames(vol)

# Use adjusted p-value; protect against zeros for -log10
vol$adj.P.Val[vol$adj.P.Val == 0] <- .Machine$double.xmin

vol$Sig <- "NotSig"
vol$Sig[vol$logFC >  lfc_cut & vol$P.Value < p_cut] <- "Up"
vol$Sig[vol$logFC < -lfc_cut & vol$P.Value < p_cut] <- "Down"

# Pick top genes to label (top by FDR within Up/Down)
top_n <- 10  # change to 5–15 as you like

top_up <- vol[vol$Sig == "Up", ]
top_up <- top_up[order(top_up$P.Value), , drop=FALSE]
top_up <- head(top_up, min(top_n, nrow(top_up)))

top_dn <- vol[vol$Sig == "Down", ]
top_dn <- top_dn[order(top_dn$P.Value), , drop=FALSE]
top_dn <- head(top_dn, min(top_n, nrow(top_dn)))

label_df <- rbind(top_up, top_dn)

# Volcano plot
p_vol <- ggplot(vol, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Sig), alpha = 0.7, size = 1.6) +
  geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
  geom_hline(yintercept = -log10(p_cut), linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = Gene),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.35,
    point.padding = 0.2
  ) +
  theme_bw() +
  labs(
    title = "Volcano Plot: AD vs Control",
    x = "log2 Fold Change (AD vs Control)",
    y = "-log10(FDR)"
  )

print(p_vol)
ggsave(result_file("_volcano_AD_vs_Control.png"), p_vol, width = 8, height = 6, dpi = 300)

# Quick counts
cat("Significant Up:", sum(vol$Sig == "Up"), "\n")
cat("Significant Down:", sum(vol$Sig == "Down"), "\n")
```

## Session information

```{r session_info}
writeLines(capture.output(sessionInfo()), result_file("sessionInfo.txt"))
sessionInfo()
```
