---
title: "Differential Gene Expression Analysis: AD vs. Control (GSE203206)"
author: "Analysis Pipeline"
date: "2025-08-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This document outlines the steps for performing a differential gene expression (DGE) analysis using RNA-Seq count data. The goal is to identify genes that are differentially expressed between Alzheimer's Disease (AD) samples and Control samples from the **GSE203206** dataset. The analysis pipeline uses the **limma-voom** workflow, which is well-suited for this type of data.

The main steps are:
1.  Data loading and preparation.
2.  Data visualization and filtering.
3.  Differential expression analysis using `limma` and `voom`.
4.  Saving the results.

***

## 1. Environment Setup

### A. Project-relative paths

All inputs and outputs are resolved from the repository root with `here::here()`.

```{r paths}
dataset_id <- "GSE203206"
input_file <- here::here("data", "processed", dataset_id, "GSE203206_Subramaniam.ADRC_brain.counts.tsv")
outdir <- here::here("results", "differential_expression", dataset_id)
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
result_file <- function(filename) file.path(outdir, filename)
```

### B. Load Required Libraries

We load the R packages required for the analysis.
* **edgeR**: Used for creating the `DGEList` object, filtering, and normalization.
* **limma**: The core package for differential expression analysis using linear models, including the `voom` function.
* **ggplot2**: A powerful library for creating visualizations, such as boxplots.
* **reshape2**: Used for data manipulation, specifically the `melt` function to prepare data for `ggplot2`.

```{r load_libraries}
library(edgeR)
library(limma)
library(ggplot2)
library(reshape2)
library(here)
```

***

## 2. Data Loading and Preparation

### A. Load Raw Count Data

We load the raw gene counts from a tab-separated values (TSV) file using `read.delim`.

```{r load_data}
counts_raw <- read.delim(input_file, check.names = FALSE, stringsAsFactors = FALSE)
```

### B. Clean, Select, and Prepare Data

The raw data is processed to prepare it for analysis.
1.  **Format and Clean Data**: We set the first column ("Gene") as the row names. Any missing values (`NA`) in the dataset are replaced with 0.
2.  **Select Samples**: We specifically select the columns corresponding to AD samples (names starting with "E" or "L") and Control samples (names starting with "C").
3.  **Create Numeric Matrix**: The final subset of data is converted into a numeric matrix, which is the required input format for `edgeR`.

```{r clean_select_counts}
# Assume first column is gene symbol/ID and set it as rownames
colnames(counts_raw)[1] <- "Gene"
rownames(counts_raw) <- counts_raw$Gene

# Replace any NA values with 0
counts_raw[is.na(counts_raw)] <- 0

# Select AD (E* or L*) and Control (C*) sample columns
all_cols <- colnames(counts_raw)
ad_cols  <- all_cols[grepl("^(E|L)", all_cols)]
ctl_cols <- all_cols[grepl("^C", all_cols)]

# Keep only the selected columns
counts_df <- counts_raw[, c("Gene", ad_cols, ctl_cols), drop = FALSE]

# Build the final numeric counts matrix
counts_mat <- counts_df[, -1, drop = FALSE]
counts_mat <- as.matrix(sapply(counts_mat, as.numeric))
rownames(counts_mat) <- counts_df$Gene
```

### C. Create Metadata

We create a metadata data frame for the selected samples. The condition ("AD" or "Control") is determined from the sample name prefixes. This metadata is then aligned with the columns of our count matrix.

```{r create_metadata}
sample_names <- colnames(counts_mat)
condition <- ifelse(grepl("^(E|L)", sample_names), "AD",
                    ifelse(grepl("^C", sample_names), "Control", NA))

metadata <- data.frame(Sample = sample_names,
                       Condition = factor(condition, levels = c("Control", "AD")))

# Align metadata and count matrix
keep_samps <- !is.na(metadata$Condition)
metadata   <- metadata[keep_samps, , drop = FALSE]
counts_mat <- counts_mat[, metadata$Sample, drop = FALSE]

cat("Samples per group:\n")
print(table(metadata$Condition))
```

***

## 3. Data Quality Assessment and Filtering

### A. Boxplot of Raw Counts (Before Filtering)

To assess the distribution of counts across samples, we generate a boxplot. This helps identify any major outliers or batch effects. We use log2-transformed counts (`log2(count + 1)`) for better visualization.

```{r boxplot_before}
log_counts <- log2(counts_mat + 1)
log_counts_df <- as.data.frame(log_counts)
log_counts_df$Gene <- rownames(log_counts_df)
melted_counts <- melt(log_counts_df, id.vars = "Gene",
                      variable.name = "Sample", value.name = "LogCount")

ggplot(melted_counts, aes(x = Sample, y = LogCount)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(Count + 1)") +
  ggtitle("Boxplot: Raw Log-Counts (AD vs Control)")

ggsave(result_file("_boxplot_raw_logcounts.png"), width = 10, height = 5, dpi = 300)
```

### B. Create DGEList, Normalize, and Filter

Now, we prepare the data for differential expression analysis.
1.  **DGEList Object**: We create a `DGEList` object, the main data container used by `edgeR`.
2.  **Normalization**: We apply TMM (Trimmed Mean of M-values) normalization to adjust for differences in library size between samples.
3.  **Filtering**: We filter out genes with low expression that provide little statistical power. We keep genes with a log2-CPM value greater than 1 in at least the minimum number of samples in any condition.

```{r filter_normalize}
dge <- DGEList(counts = counts_mat, samples = metadata, group = metadata$Condition)
dge <- calcNormFactors(dge, method = "TMM")
cat(sprintf("Genes before filtering: %d\n", nrow(dge)))

# Define filtering threshold
log2cpm_raw <- cpm(dge, log = TRUE, prior.count = 1)
min_n <- min(table(metadata$Condition))
keep <- rowSums(log2cpm_raw > 1) >= min_n

# Filter the DGEList and recalculate norm factors
dge_filt <- dge[keep, , keep.lib.sizes = FALSE]
dge_filt <- calcNormFactors(dge_filt, method = "TMM")

cat(sprintf("Kept genes after filtering: %d\n", nrow(dge_filt)))
```

### C. Boxplot of Filtered Counts (After Filtering)

We create another boxplot to visualize the data after filtering and normalization. The distributions between samples should now be more aligned.

```{r boxplot_after}
log2cpm_filtered <- cpm(dge_filt, log = TRUE, prior.count = 1)
log_counts_df_after <- as.data.frame(log2cpm_filtered)
log_counts_df_after$Gene <- rownames(log_counts_df_after)
melted_counts_after <- melt(log_counts_df_after, id.vars = "Gene",
                            variable.name = "Sample", value.name = "Log2CPM")

ggplot(melted_counts_after, aes(x = Sample, y = Log2CPM)) +
  geom_boxplot(outlier.size = 0.3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ylab("Log2(CPM)") +
  ggtitle("Boxplot: Filtered Log-CPM (AD vs Control)")

ggsave(result_file("_boxplot_filtered_logcpm.png"), width = 10, height = 5, dpi = 300)

write.csv(log2cpm_filtered,result_file("GSE203206_Expression_data.csv"))
```

***

## 4. Differential Expression Analysis with limma-voom

### A. Define the Design Matrix

We create a design matrix that specifies the model to be fit to each gene. The formula `~ Condition` models the gene expression based on the `Condition` variable (AD vs. Control), with "Control" as the reference level.

```{r design_matrix}
design <- model.matrix(~ Condition, data = dge_filt$samples)
rownames(design) <- colnames(dge_filt)
head(design)
```

### B. Run Voom and Fit the Linear Model

The `limma-voom` workflow transforms the count data for linear modeling.
1.  **`voom`**: Transforms the count data to log2-CPM, estimates the mean-variance relationship, and computes precision weights. The resulting trend plot is saved as a PNG.
2.  **`lmFit`**: Fits a linear model for each gene based on the design matrix.
3.  **`eBayes`**: Applies empirical Bayes moderation to the standard errors, borrowing information across genes to improve statistical power.

```{r voom_limma, fig.cap="Voom Mean-Variance Plot"}
v <- voom(dge_filt, design, plot = TRUE)
fit <- lmFit(v, design)
fit <- eBayes(fit)
```

### C. Extract and Subset DEGs

We extract the results using `topTable` and subset them to find statistically significant differentially expressed genes (DEGs). We define significance with a log-fold-change (logFC) threshold of `> 0.58` (a fold change of ~1.5) and a P-value threshold of `< 0.05`.

```{r extract_degs}
deg <- topTable(fit, coef = "ConditionAD", number = Inf, sort.by = "P")
deg_sig_adjusted <- subset(deg, abs(logFC) > 0.58 & adj.P.Val < 0.05)
deg_sig_unadjusted <- subset(deg, abs(logFC) > 0.58 & P.Value < 0.05)
# Primary filtered set for figures, gene lists, and enrichment.
deg_sig <- deg_sig_unadjusted
cat(sprintf("Found %d significant DEGs.\n", nrow(deg_sig)))
cat(sprintf("Adjusted-p-value DEGs: %d; unadjusted-p-value DEGs: %d\n",
            nrow(deg_sig_adjusted), nrow(deg_sig_unadjusted)))
```

***

## 5. Save Results

Finally, we save the lists of significant DEGs. We save several files:
1.  A CSV file with the full statistics for all significant DEGs.
2.  Text files containing only the gene names for upregulated and downregulated genes.
3.  CSV files containing the statistics for the top 20 upregulated and top 20 downregulated genes, sorted by p-value.

```{r save_results}
# Save all significant DEGs
write.csv(deg, result_file("full_DE_results.csv"), row.names = TRUE)
write.csv(deg_sig, result_file("significant_DEGs.csv"), row.names = TRUE)
write.csv(deg_sig_adjusted, result_file("significant_DEGs_adjusted.csv"), row.names = TRUE)
write.csv(deg_sig_unadjusted, result_file("significant_DEGs_unadjusted.csv"), row.names = TRUE)

# Save lists of up- and down-regulated gene names
up_genes <- rownames(deg_sig[deg_sig$logFC > 0, , drop = FALSE])
down_genes <- rownames(deg_sig[deg_sig$logFC < 0, , drop = FALSE])

write.table(up_genes, result_file("upregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)
write.table(down_genes, result_file("downregulated_genes.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)

cat(sprintf("Found %d upregulated and %d downregulated genes.\n", length(up_genes), length(down_genes)))

# Isolate and save top 20 up/down regulated genes
top20_up <- head(deg_sig[deg_sig$logFC > 0, ][order(deg_sig$P.Value[deg_sig$logFC > 0]), ], 20)
top20_down <- head(deg_sig[deg_sig$logFC < 0, ][order(deg_sig$P.Value[deg_sig$logFC < 0]), ], 20)

write.csv(top20_up, result_file("Top20_Upregulated_AD_vs_Control.csv"), row.names = TRUE)
write.csv(top20_down, result_file("Top20_Downregulated_AD_vs_Control.csv"), row.names = TRUE)
```
```{r map_deg_to_hgnc_fix}
library(AnnotationDbi)
library(org.Hs.eg.db)

map_deg_table_to_symbol <- function(tt, id_keytype = "ENSEMBL") {
  ids <- rownames(tt)
  ids_clean <- sub("\\..*$", "", ids)

  sym <- mapIds(
    org.Hs.eg.db,
    keys = ids_clean,
    keytype = id_keytype,
    column = "SYMBOL",
    multiVals = "first"
  )

  # Build a data.frame with SYMBOL as a column first (avoid duplicate rowname error)
  df <- tt
  df$SYMBOL <- unname(sym)

  # drop unmapped
  df <- df[!is.na(df$SYMBOL) & df$SYMBOL != "", , drop = FALSE]

  # keep most significant row per SYMBOL (smallest P.Value)
  df <- df[order(df$P.Value, df$adj.P.Val, na.last = TRUE), , drop = FALSE]
  df <- df[!duplicated(df$SYMBOL), , drop = FALSE]

  rownames(df) <- df$SYMBOL
  df$SYMBOL <- NULL
  df
}

deg_sym     <- map_deg_table_to_symbol(deg)
deg_sig_sym <- map_deg_table_to_symbol(deg_sig)

cat("deg rows:", nrow(deg), "->", nrow(deg_sym), "\n")
cat("deg_sig rows:", nrow(deg_sig), "->", nrow(deg_sig_sym), "\n")

```

```{r}
# -------------------------
# HEATMAP: Top 20 genes (uniform groups)
# -------------------------
suppressPackageStartupMessages({
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(matrixStats)
  library(pheatmap)
  library(clusterProfiler)
  library(enrichplot)
})

# -------------------------
# Make a SYMBOL-rowed expression matrix from voom (v$E)
# -------------------------
make_symbol_expr <- function(E, keytype = "ENSEMBL") {
  ids <- rownames(E)
  ids_clean <- sub("\\..*$", "", ids)

  sym <- mapIds(
    org.Hs.eg.db,
    keys      = ids_clean,
    keytype   = keytype,
    column    = "SYMBOL",
    multiVals = "first"
  )

  keep <- !is.na(sym) & sym != ""
  E2 <- E[keep, , drop = FALSE]
  rownames(E2) <- unname(sym[keep])

  # if multiple ENSEMBL map to same SYMBOL, keep the most variable row (best for heatmaps)
  if (any(duplicated(rownames(E2)))) {
    rv <- matrixStats::rowVars(as.matrix(E2), na.rm = TRUE)
    E2 <- E2[order(rownames(E2), -rv), , drop = FALSE]
    E2 <- E2[!duplicated(rownames(E2)), , drop = FALSE]
  }

  E2
}

E_sym <- make_symbol_expr(v$E, keytype = "ENSEMBL")

# -------------------------
# Choose top genes (HGNC SYMBOLS) from your SYMBOL-mapped DEG tables
# -------------------------
top_tbl <- if (nrow(deg_sig_sym) >= 20) deg_sig_sym else deg_sym
top20_genes <- rownames(top_tbl)[1:min(20, nrow(top_tbl))]

# Subset safely (avoids "subscript out of bounds")
genes_use <- intersect(top20_genes, rownames(E_sym))
stopifnot(length(genes_use) > 1)

expr_hm <- E_sym[genes_use, , drop = FALSE]

# Optional: keep gene order exactly as in top20_genes
expr_hm <- expr_hm[match(genes_use, rownames(expr_hm)), , drop = FALSE]
expr_hm <- expr_hm[top20_genes[top20_genes %in% rownames(expr_hm)], , drop = FALSE]

# -------------------------
# Heatmap (SYMBOLs kept)
# -------------------------
pheatmap(expr_hm, scale = "row", fontsize_row = 9)


# Order samples: Control then AD
ord <- order(metadata$Condition)   # factor levels are Control, AD (as you set)
expr_hm <- expr_hm[, ord, drop = FALSE]
meta_hm <- metadata[ord, , drop = FALSE]

ann_col <- data.frame(Condition = meta_hm$Condition)
rownames(ann_col) <- meta_hm$Sample
ann_colors <- list(Condition = c(Control = "#3B6FB6", AD = "#C0392B"))

# add a visual gap between groups
n_ctl <- sum(meta_hm$Condition == "Control")
gaps <- n_ctl

png(result_file("_heatmap_top20_genes.png"), width = 4000, height = 2400, res = 600)
pheatmap(
  expr_hm,
  scale = "row",
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = FALSE,
  fontsize_row = 9,
  cluster_cols = FALSE,   # keeps Control together, AD together
  gaps_col = gaps,
  main = "Heatmap Top 20 genes"
)
dev.off()
```

```{r}
# assumes rownames are gene SYMBOLs; if they are ENSG IDs, tell me and I'll swap mapping

sig_symbols <- rownames(deg_sig)

conv <- bitr(sig_symbols,
fromType = "ENSEMBL",
toType   = c("ENTREZID"),
OrgDb    = org.Hs.eg.db)

# keep unique ENTREZ

sig_entrez <- unique(conv$ENTREZID)

# up/down sets (optional separate enrichment)

up_symbols   <- rownames(deg_sig[deg_sig$logFC > 0, , drop=FALSE])
down_symbols <- rownames(deg_sig[deg_sig$logFC < 0, , drop=FALSE])

conv_up <- bitr(up_symbols, fromType="ENSEMBL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
conv_dn <- bitr(down_symbols, fromType="ENSEMBL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

up_entrez <- unique(conv_up$ENTREZID)
dn_entrez <- unique(conv_dn$ENTREZID)

cat("Mapped significant genes to Entrez:", length(sig_entrez), "\n")
cat("Up:", length(up_entrez), "Down:", length(dn_entrez), "\n")
```

```{r}
ego_bp <- enrichGO(gene          = up_entrez,
OrgDb         = org.Hs.eg.db,
keyType       = "ENTREZID",
ont           = "BP",
pAdjustMethod = "BH",
qvalueCutoff  = 0.05,
readable      = TRUE)

ego_cc <- enrichGO(gene = up_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="CC", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

ego_mf <- enrichGO(gene = up_entrez, OrgDb=org.Hs.eg.db, keyType="ENTREZID",
ont="MF", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)

# Save tables

write.csv(as.data.frame(ego_bp), result_file("GSE203206_GO_BP_UP.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_cc), result_file("GSE203206_GO_CC_UP.csv"), row.names = FALSE)
write.csv(as.data.frame(ego_mf), result_file("GSE203206_GO_MF_UP.csv"), row.names = FALSE)

# Plot

p1 <- dotplot(ego_bp, showCategory = 20) + ggtitle("GO Biological Process (ORA)")
p2 <- dotplot(ego_cc, showCategory = 15) + ggtitle("GO Cellular Component (ORA)")
p3 <- dotplot(ego_mf, showCategory = 15) + ggtitle("GO Molecular Function (ORA)")

print(p1); print(p2); print(p3)

ggsave(result_file("_GO_BP_dotplot.png"), p1, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_CC_dotplot.png"), p2, width = 9, height = 6, dpi = 300)
ggsave(result_file("_GO_MF_dotplot.png"), p3, width = 9, height = 6, dpi = 300)
```

```{r}
ekegg_up <- enrichKEGG(gene=up_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

ekegg_dn <- enrichKEGG(gene=dn_entrez, organism="hsa",
                       pAdjustMethod="BH", qvalueCutoff=0.05)

# Convert Entrez IDs in results to gene symbols for readability
ekegg_up <- setReadable(ekegg_up, OrgDb=org.Hs.eg.db, keyType="ENTREZID")
ekegg_dn <- setReadable(ekegg_dn, OrgDb=org.Hs.eg.db, keyType="ENTREZID")

write.csv(as.data.frame(ekegg_up), result_file("KEGG_UP.csv"), row.names=FALSE)
write.csv(as.data.frame(ekegg_dn), result_file("KEGG_DOWN.csv"), row.names=FALSE)

p_kegg_up <- dotplot(ekegg_up, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Upregulated)")
p_kegg_dn <- dotplot(ekegg_dn, showCategory=20) + ggtitle("KEGG Pathway Enrichment (Downregulated)")

print(p_kegg_up); print(p_kegg_dn)

ggsave(result_file("_KEGG_up_dotplot.png"), p_kegg_up, width=9, height=10, dpi=300)
ggsave(result_file("_KEGG_down_dotplot.png"), p_kegg_dn, width=9, height=10, dpi=300)

```

```{r}
library(ggplot2)
library(ggrepel)

# thresholds (match your DEG definition)
lfc_cut <- 0.58
p_cut   <- 0.05   # using P.Value only

# Use SYMBOL-mapped significant table directly
vol <- deg_sig_sym
vol$Gene <- rownames(vol)   # HGNC symbols

# protect against zeros for -log10
vol$P.Value[vol$P.Value == 0] <- .Machine$double.xmin

# classify using P.Value (no FDR anywhere)
vol$Sig <- "NotSig"
vol$Sig[vol$logFC >  lfc_cut & vol$P.Value < p_cut] <- "Up"
vol$Sig[vol$logFC < -lfc_cut & vol$P.Value < p_cut] <- "Down"

# Pick top genes to label (smallest P.Value within Up/Down)
top_n <- 10

top_up <- vol[vol$Sig == "Up", , drop = FALSE]
top_up <- top_up[order(top_up$P.Value), , drop = FALSE]
top_up <- head(top_up, min(top_n, nrow(top_up)))

top_dn <- vol[vol$Sig == "Down", , drop = FALSE]
top_dn <- top_dn[order(top_dn$P.Value), , drop = FALSE]
top_dn <- head(top_dn, min(top_n, nrow(top_dn)))

label_df <- rbind(top_up, top_dn)

# Volcano plot (HGNC labels; p-value axis)
p_vol <- ggplot(vol, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Sig), alpha = 0.7, size = 1.6) +
  geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
  geom_hline(yintercept = -log10(p_cut), linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = Gene),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.35,
    point.padding = 0.2
  ) +
  theme_bw() +
  labs(
    title = "Volcano Plot: AD vs Control (HGNC symbols)",
    x = "log2 Fold Change (AD vs Control)",
    y = "-log10(p-value)"
  )

print(p_vol)
ggsave(result_file("_volcano_AD_vs_Control_pvalue_deg_sig_sym.png"), p_vol, width = 8, height = 6, dpi = 300)

cat("Up:", sum(vol$Sig == "Up"), "\n")
cat("Down:", sum(vol$Sig == "Down"), "\n")
```

## Session information

```{r session_info}
writeLines(capture.output(sessionInfo()), result_file("sessionInfo.txt"))
sessionInfo()
```
```
```
