---
title: "Differential Gene Expression Analysis: AD vs. Control by region (GSE278723)"
author: "KASHVI SHAH"
date: "`2025-08-19`"
Author: 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  region_files:
    - "output_CA1.csv"
    - "output_CA2.csv"
    - "output_CA3.csv"
    - "output_CA4.csv"
    - "output_NX.csv"
  logFC_cut: 0.58       # ~1.5x fold change
  p_cut: 0.05           # retained from original analysis
---

# 1.Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(edgeR)
  library(limma)
  library(ggplot2)
  library(reshape2)
  library(here)
})

# Parameters. Knitting supplies `params`; interactive chunk execution does not,
# so use the YAML defaults when no parameter list is present.
analysis_params <- get0("params", ifnotfound = NULL, inherits = TRUE)
if (!is.list(analysis_params)) {
  analysis_params <- list(
    region_files = c(
      "output_CA1.csv", "output_CA2.csv", "output_CA3.csv",
      "output_CA4.csv", "output_NX.csv"
    ),
    logFC_cut = 0.58,
    p_cut = 0.05
  )
}

dataset_id   <- "GSE278723"
input_dir    <- here::here("data", "processed", dataset_id)
region_files <- file.path(input_dir, analysis_params$region_files)
logFC_cut    <- analysis_params$logFC_cut
p_cut        <- analysis_params$p_cut
outdir_root  <- here::here("results", "differential_expression", dataset_id)

# Output root
dir.create(outdir_root, showWarnings = FALSE, recursive = TRUE)

# Helper theme
theme_set(theme_bw(base_size = 11))
```

# 2.Methods (brief)

  Read per-region count CSV (first column = gene; remaining columns = samples).

  Basic QC: duplicates/NA report; log2(count+1) boxplot.

  edgeR: TMM normalization, gene filtering by CPM.

  voom/limma: linear model with Condition (Control reference), empirical Bayes.

  Primary filtered analysis below: |logFC| > `r logFC_cut` and nominal P.Value < `r p_cut`.

  Save: full table, significant DEGs, up/down gene lists, top-20 by p-value, and voom       mean–variance plot.

  The unadjusted DEG set is the primary filtered set used for downstream outputs;
  adjusted and complete unfiltered result tables are also saved separately.

# 3.Region Analysis Function

```{r}
analyze_one <- function(infile) {
  tag <- sub("^output_(.*)\\.csv$", "\\1", basename(infile))   # e.g., "CA1"
  outdir <- file.path(outdir_root, tag)
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

  message("\n========== Running: ", tag, " ==========")

  # --- Read counts ---
  counts_raw <- read.csv(infile, check.names = FALSE, stringsAsFactors = FALSE)

  # Ensure first column is "Gene" + clean colnames (remove leading X)
  colnames(counts_raw)[1] <- "Gene"
  colnames(counts_raw) <- sub("^X", "", colnames(counts_raw))

  # --- Pre-checks (report, then fix only if needed) ---
  dup_n <- sum(duplicated(counts_raw[,1]))
  na_n  <- sum(is.na(counts_raw))
  message("[", tag, "] Duplicated genes: ", dup_n, " | Total NA values: ", na_n)

  if (dup_n > 0) counts_raw$Gene <- make.unique(as.character(counts_raw$Gene)) else counts_raw$Gene <- as.character(counts_raw$Gene)
  rownames(counts_raw) <- counts_raw$Gene
  if (na_n > 0) counts_raw[is.na(counts_raw)] <- 0

  # Keep Gene first, then all sample columns
  counts_df <- counts_raw[, c("Gene", setdiff(colnames(counts_raw), "Gene")), drop = FALSE]

  # Build numeric counts matrix
  counts_mat <- counts_df[, -1, drop = FALSE]
  counts_mat <- as.matrix(sapply(counts_mat, as.numeric))
  rownames(counts_mat) <- counts_df$Gene

  # --- Metadata (set per region; edit counts to match your files) ---
  sample_names <- colnames(counts_mat)

  # Group labels come from the donor ID embedded in each column name
  # (e.g. "14.CA2" -> donor "14"), joined to Supplementary Table 1 of the source
  # paper (Li C et al. 2024, Aging Dis), which records the diagnostic group for
  # all 50 donors: NC 01-13 (n = 13) and AD 14-28 (n = 15).
  #
  # The join is on donor ID rather than column position, so subregions with
  # donors missing are handled correctly (CA1 does not contain every donor).
  donors <- read.csv(
    here::here("data", "metadata", "GSE278723",
               "GSE278723_supplementary_table1_donors.csv"),
    colClasses = c(Donor = "character"), stringsAsFactors = FALSE)

  donor_id <- sub("\\..*$", "", sample_names)
  grp <- donors$Group[match(donor_id, donors$Donor)]
  if (anyNA(grp)) {
    stop(sprintf("[%s] No Supplementary Table 1 entry for donor(s): %s",
                 tag, paste(unique(donor_id[is.na(grp)]), collapse = ", ")))
  }
  # Only the NC and AD donors are retained; CAA and AD+CAA are out of scope.
  condition <- ifelse(grp == "NC", "Control", ifelse(grp == "AD", "AD", NA))
  if (anyNA(condition)) {
    message("[", tag, "] Excluding ", sum(is.na(condition)),
            " sample(s) from groups other than NC/AD: ",
            paste(unique(grp[is.na(condition)]), collapse = ", "))
  }

  metadata <- data.frame(Sample = sample_names,
                         Condition = factor(condition, levels = c("Control","AD")))
  keep_samps <- !is.na(metadata$Condition)
  if (any(!keep_samps)) {
    message("[", tag, "] Dropping ", sum(!keep_samps), " samples with unknown group labels.")
  }
  metadata   <- metadata[keep_samps, , drop = FALSE]
  counts_mat <- counts_mat[, metadata$Sample, drop = FALSE]

  message("[", tag, "] Samples per group:")
  print(table(metadata$Condition))

  # --- QC Boxplot (log2 count) ---
  log_counts <- log2(counts_mat + 1)
  log_counts_df <- as.data.frame(log_counts)
  log_counts_df$Gene <- rownames(log_counts_df)
  melted_counts <- reshape2::melt(log_counts_df, id.vars = "Gene",
                                  variable.name = "Sample", value.name = "LogCount")

  p <- ggplot(melted_counts, aes(x = Sample, y = LogCount)) +
    geom_boxplot(outlier.size = 0.3) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
    ylab("Log2(Count + 1)") +
    ggtitle(sprintf("%s: Boxplot Raw Counts (AD vs Control)", tag))
  
  print(p)
  ggsave(file.path(outdir, sprintf("%s_boxplot_raw_logcounts.png", tag)),
         p, width = 10, height = 5, dpi = 300)

  # --- edgeR: DGEList, TMM, Filter ---
  dge <- DGEList(counts = counts_mat, samples = metadata, group = metadata$Condition)
  dge <- calcNormFactors(dge, method = "TMM")

  log2cpm_raw <- cpm(dge, log = TRUE, prior.count = 1)
  min_n <- min(table(metadata$Condition))
  keep <- rowSums(log2cpm_raw > 1) >= min_n

  dge_filt <- dge[keep, , keep.lib.sizes = FALSE]
  dge_filt <- calcNormFactors(dge_filt, method = "TMM")

  message("[", tag, "] Kept genes after filtering: ", nrow(dge_filt))
  
  # Recalculate the log2-CPM on the filtered DGEList object.
  log2cpm_filtered <- cpm(dge_filt, log = TRUE, prior.count = 1)

  log_counts_df_after <- as.data.frame(log2cpm_filtered)
  log_counts_df_after$Gene <- rownames(log_counts_df_after)
  melted_counts_after <- melt(log_counts_df_after, id.vars = "Gene",
                            variable.name = "Sample", value.name = "LogCount")

  p_after <- ggplot(melted_counts_after, aes(x = Sample, y = LogCount)) +
    geom_boxplot(outlier.size = 0.3) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
    ylab("Log2(CPM + 1)") +
    ggtitle("Boxplot of Filtered & Normalized Log-CPM")
  
  print(p_after)

  ggsave(file.path(outdir, sprintf("%s_boxplot_filtered_logcpm.png", tag)), 
         p_after, width = 10, height = 5, dpi = 300)
  
  write.csv(log2cpm_filtered,
  file = file.path(outdir, sprintf("%s_filtered_logcpm.csv", tag)),
  row.names = TRUE)
  
  # --- Design (Control reference) ---
  design <- model.matrix(~ Condition, data = dge_filt$samples)

  # --- Voom + Limma ---
  v <- voom(dge_filt, design, plot = TRUE)
  fit <- lmFit(v, design)
  fit <- eBayes(fit)

  deg <- topTable(fit, coef = "ConditionAD", number = Inf, sort.by = "P")

  # --- Significant DEGs (effect size + p) ---
  deg_sig_adjusted <- subset(deg, abs(logFC) > logFC_cut & adj.P.Val < p_cut)
  deg_sig_unadjusted <- subset(deg, abs(logFC) > logFC_cut & P.Value < p_cut)
  # Primary filtered set for figures, gene lists, and enrichment.
  deg_sig <- deg_sig_unadjusted

  message("[", tag, "] Differentially expressed genes:", nrow(deg_sig))
  message("[", tag, "] Adjusted-p-value DEGs: ", nrow(deg_sig_adjusted),
          "; unadjusted-p-value DEGs: ", nrow(deg_sig_unadjusted))
  
  # --- Save outputs ---
  write.csv(deg,     file.path(outdir, "full_DE_results.csv"), row.names = TRUE)
  write.csv(deg_sig, file.path(outdir, "significant_DEGs.csv"), row.names = TRUE)
  write.csv(deg_sig_adjusted,
            file.path(outdir, "significant_DEGs_adjusted.csv"), row.names = TRUE)
  write.csv(deg_sig_unadjusted,
            file.path(outdir, "significant_DEGs_unadjusted.csv"), row.names = TRUE)

  up_genes   <- rownames(deg_sig[deg_sig$logFC > 0, , drop = FALSE])
  
  message("[", tag, "] Upregulated genes:", length(up_genes))
  
  down_genes <- rownames(deg_sig[deg_sig$logFC < 0, , drop = FALSE])
  
  message("[", tag, "] Downregulated genes:", length(down_genes))

  write.table(up_genes,   file.path(outdir, "upregulated_genes.txt"),
              quote = FALSE, row.names = FALSE, col.names = FALSE)
  write.table(down_genes, file.path(outdir, "downregulated_genes.txt"),
              quote = FALSE, row.names = FALSE, col.names = FALSE)

  up_tbl   <- deg_sig[deg_sig$logFC > 0, , drop = FALSE]
  down_tbl <- deg_sig[deg_sig$logFC < 0, , drop = FALSE]
  top20_up   <- head(up_tbl[order(up_tbl$P.Value), ], 20)
  top20_down <- head(down_tbl[order(down_tbl$P.Value), ], 20)

  write.csv(top20_up,   file.path(outdir, sprintf("%s_Top20_Upregulated_AD_vs_Control.csv", tag)))
  write.csv(top20_down, file.path(outdir, sprintf("%s_Top20_Downregulated_AD_vs_Control.csv", tag)))

  message("========== Done: ", tag, " ==========")

  # Return invisibly for inline summaries
  invisible(list(
    tag = tag,
    n_sig = nrow(deg_sig),
    n_up = nrow(up_tbl),
    n_down = nrow(down_tbl)
  ))
}
```

# 4.Run All Regions
```{r}
for (f in region_files) analyze_one(f)
```

# 5.Session information

```{r session_info}
writeLines(capture.output(sessionInfo()), file.path(outdir_root, "sessionInfo.txt"))
sessionInfo()
```
